\documentclass{article}
\usepackage{blindtext}
\usepackage{geometry}
\usepackage{graphicx}
\graphicspath{{./figs/}}
\usepackage{multirow, multicol, longtable}
\usepackage{url}
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue]{hyperref}
\geometry{
	a4paper,
	total={170mm,257mm},
	left=20mm,
	top=20mm,
}
\usepackage[authoryear]{natbib} 
\usepackage{longtable}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{anyfontsize}
\usepackage{lineno}
\usepackage{authblk}
\linenumbers
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{amsmath, amsfonts, booktabs, textcomp}

\title{
	A High-Performance Rule-Based Multi-Agent Framework for Spatio-Temporal 
	Vector Dynamics: Application to \textit{Anopheles stephensi} in Ethiopia}

\author[1,2]{Salvador D. Atagong}
\author[1]{Komi Mensah Agboka}
\author[1]{Kennedy Senagi}
\author[2,3]{Henri Tonnang}
\author[2]{John Odindi}

\affil[1]{International Center of Insect Physiology and Ecology (ICIPE)}
\affil[2]{University of KwaZulu-Natal, South Africa}
\affil[3]{International Institute of Tropical Agriculture (IITA), Ibadan, Nigeria}
\date{}

\begin{document}
\maketitle

\begin{abstract}


\paragraph{Keywords: multi-agent systems, vector-borne disease, \textit{Anopheles stephensi}, Metzler matrix, 
high-performance computing, spatial ecology.} 
\end{abstract}

\section{Introduction}
\label{sec:introduction}
\par
Vector-borne diseases (VBDs) represent one of the most profound challenges to global health security, accounting for 
approximately 700,000 deaths annually and imposing significant economic burdens on developing nations \citep{2025.RL.HMD}. 
Malaria alone threatens nearly half the world's population, with Sub-Saharan Africa bearing 94\% of the global case burden 
as \citep{2024.WH.WWM}. While the past two decades witnessed substantial reductions in malaria mortality through coordinated 
interventions, this hard-won progress faces unprecedented threats from ecological and demographic shifts. 
The rapid urbanization of Sub-Saharan Africa, currently the world's fastest urbanizing region \citep{2019.IS.UAI}, 
has created novel transmission ecologies that challenge historical assumptions about vector ecology and control 
strategies \citep{2024.NK.MBD}.
\par
The primary driver of this epidemiological transition is the globalization of previously localized vectors, particularly 
the invasive malaria mosquito \textit{Anopheles stephensi}. Unlike endemic African malaria vectors that predominantly 
breed in natural water bodies, \textit{An. stephensi} exhibits remarkable ecological plasticity, thriving in artificial 
containers and man-made structures characteristic of urban environments \citep{2025.GC.SAB}. This species shares critical 
adaptive traits with \textit{Aedes aegypti}, the primary dengue vector, including container breeding, anthropophily, and 
exophilic resting behaviors \citep{2022.AH.TPI}. The consequences of this adaptation were discored in Djibouti, 
where\textit{ An. stephensi} establishment triggered a 30-fold increase in malaria cases within seven years, transforming a 
nation on the verge of malaria elimination into a high-transmission zone \citep{2022.AH.TPI}. This vector's capacity to exploit 
previously low-risk urban settings necessitates fundamentally new approaches to surveillance, modeling, and 
intervention design \citep{2024.TM.AQA}.
\par
A critical bottleneck in responding to this emerging threat is the pervasive "Data Gap" characterizing many resource-constrained 
settings where \textit{An. stephensi} is establishing. Effective vector control relies on robust decision-support systems built 
upon high-quality, high-resolution entomological and epidemiological data, yet national surveillance programs often operate with 
limited financial and technological resources \citep{2025.RL.HMD}. Current modeling approaches struggle to bridge the gap between 
theoretical elegance and practical utility: empirical models propagate measurement errors through unreliable data streams, while 
process-based models frequently lack the spatial granularity and computational efficiency needed for operational 
deployment \citep{2025.RL.HMD}. This disconnect is particularly acute for invasive vectors, where absence of historical 
baseline data undermines traditional statistical approaches \citep{2019.EL.ALO}.
\par
To address these challenges, we propose a novel computational framework that integrates high-performance computing methodologies 
with ecological process modeling. Our system implements a layered, rule-based Multi-Agent Simulation (MAS) architecture that 
harmonizes heterogeneous spatio-temporal data streams into a unified analytical environment. This framework is part of the 
family of ABS (Agent Based Simulation) which are simulation models where the researcher explicitly describe the decision 
processes of simulated actors at micro-level \citep{2008.PE.ITM}. The proposed framework bridges Computer Science and 
Environmental Science by conceptualizing environmental data as dynamic information layers, including static rasters (for instance: 
elevation, building density, population density), time-series climate variables, and vector data (shapefiles), that interact with 
autonomous agent populations. We distinguish between two agent classes: "Inert Agents" representing static environmental features 
(urban water containers serving as breeding sites) and "Living Agents" modeling biological vectors (An. stephensi mosquitoes for 
example) with behavioral rules derived from empirical studies.
\par
This study demonstrates how such synthetic modeling approaches can provide critical insights where field data are scarce or 
unreliable. Using Dire Dawa, and Somali woredas from Ethiopia, which are regions experiencing \textit{An. stephensi} invasion, 
\citep{2024.GZ.ROC,2023.DH.FRO} as a proof-of-concept, we simulate 6 months of mosquito population dynamics across the whole regions. 
Our results illustrate how agent-based modeling can identify high-risk micro-habitats, simulate seasonal population trajectories, and 
potentially assist intervention strategies in data-limited settings. By making the simulation framework open-source and computationally 
efficient, we aim to empower public health agencies in resource-constrained regions with decision-support tools capable of addressing the 
unique challenges posed by urban-adapted invasive vectors.


\section{Materials and Methods}
\label{sec:methods}
\subsection{Framework Architecture}
\label{sec:framework}
\subsubsection{Overview}
% comprehensive chart of the framework
The simulation framework follows a modular, agent-based design that integrates spatially explicit environmental data, 
temperature-dependent life-cycle models, and behavioural rules. It is implemented in Java and built around four core 
subsystems: (i) a time manager that advances simulation time in discrete ticks (15-minutes per tick) and interpolates 
climate data; (ii) a spatial registry that indexes agents on a lazy grid for efficient neighbourhood queries; 
(iii) an agent layer system that manages agents and evaluates rules in parallel; and (iv) a lifecycle model that computes 
temperature-driven development and survival rates. The overall architecture is illustrated in Figure~\ref{fig:arch}.
\begin{figure}[htbp]
\centering
\begin{minipage}{0.9\textwidth}
\begin{verbatim}
<!-- Mermaid chart: framework architecture -->
graph TD
    A[TimeManager] --> B[SimulationEngine]
    B --> C[SpatialRegistry]
    B --> D[AgentLayer: Mosquitoes]
    B --> E[AgentLayer: WaterTanks]
    D --> F[RuleEngine]
    E --> F
    D --> G[LifecycleModel]
    B --> H[Raster Layers (Climate, Static)]
    B --> I[Export Service]
    I --> J[CSV Snapshots]
    J --> K[Analysis Notebook]
\end{verbatim}
\caption{High-level architecture of the simulation framework.}
\label{fig:arch}
\end{minipage}
\end{figure}

\subsubsection{Core Components}
% one fleshy paragraph per components (agent layers, raster layers, lifecycle models, etc.)
\paragraph{Agent layers.} Two agent types are implemented: \texttt{LivingAgent} (representing a mosquitoe) 
and \texttt{InertAgent} (representing a breeding site). Living agents track life stage (egg, larva, pupa, adult), age, 
energy, gravid status, and resting behaviour. Their movement is modelled as a random walk with a step size derived from field observations 
of flight range \citep{1966.MA.FRL}. Inert agents maintain water volume, egg count, larval count, and carrying capacity; they update their 
capacity dynamically based on water level, mimicking the density-dependent mortality observed in larval habitats \citep{2016.ST.OTI}.  
Agent layers are processed in parallel using a partition-based container (\texttt{AgentContainer}) and a rule-engine that evaluates 
behavioural rules (e.g., feeding, oviposition, resting) in priority order \citep{2023.ST.ASR}.

\paragraph{Raster layers and climate data.}  
Static raster layers (elevation, building density, population density) are loaded from GeoTIFF files using GeoTools and stored either 
in memory or memory-mapped for large grids \citep{2021.MB.AUO}. Climate time series (2-m temperature and total precipitation) are ingested 
from NetCDF files (ERA5-Land reanalysis) into interpolated raster layers that perform linear temporal interpolation between hourly 
steps \citep{2014.GC.AED, 2017.CL.IMS}.

\paragraph{Lifecycle model.} 
The lifecycle model represents mosquito population dynamics using a discrete-time, stage-structured Metzler matrix 
\citep{2023.ST.ASR, 2025.KM.WTR}. Four developmental stages are tracked: eggs ($E$), larvae ($L$), pupae ($P$), and adults ($A$). 
All rates are temperature-dependent and defined on a per-day basis, then converted to tick-level probabilities via the Poisson 
process ($p = 1 - \exp(-r \Delta t)$, where $\Delta t$ is the tick duration in days). The daily projection matrix is:

\[
\mathbf{M}(T) = 
\begin{pmatrix}
(1 - d_E) S_E & 0 & 0 & F \cdot K \\
d_E S_E & (1 - d_L) S_L & 0 & 0 \\
0 & d_L S_L & (1 - d_P) S_P & 0 \\
0 & 0 & d_P S_P & 1 - \mu_A
\end{pmatrix}
\]

with stage abundances evolving as $[E,L,P,A]_{t+1}^\top = \mathbf{M}(T) [E,L,P,A]_t^\top$. The temperature-dependent functions are:

\begin{itemize}
  \item \textbf{Development rates} (day$^{-1}$): $d_X(T) = \max(0, a_X T^2 + b_X T + c_X)$ for $X \in \{E,L,P\}$.
  \item \textbf{Stage survival} (Gaussian): $S_X(T) = \text{amp}_X \cdot \exp\left(-\frac{1}{2}\left(\frac{T - \mu_X}{\sigma_X}\right)^2\right)$.
  \item \textbf{Adult mortality} (day$^{-1}$): $\mu_A(T) = \max(0, a_A T^2 + b_A T + c_A)$.
  \item \textbf{Fecundity} (eggs/female/day): $F(T) = \max\left(0, a_F e^{b_F T} - e^{b_F T_{\max} - \left(\frac{T_{\max}-T}{c_F}\right)^2}\right)$.
  \item \textbf{Host carrying capacity}: $K = 1 + \ln(1 + \text{population density})$, which scales fecundity based on human population density.
  \item \textbf{Density-dependent larval mortality}: each aquatic habitat has a carrying capacity $\text{cap}$ (function of water volume); 
  if larval count exceeds $\text{cap}$, excess larvae are removed probabilistically.
\end{itemize}

All temperature-response coefficients are stored in a YAML configuration file under the \texttt{species} key, making the model generic and easily 
adaptable to any mosquito species by replacing the coefficients in Table~\ref{tab:lifecycle_params}. This parametrisation 
approach follows the principles of generic insect population modelling \citep{2025.KM.DPF, 2025.KM.WTR, 2026.KM.THM}.

\begin{table}[htbp]
\centering
\caption{Generic temperature-response parameters for a mosquito lifecycle model.}
\label{tab:lifecycle_params}
\begin{tabular}{llll}
\toprule
Parameter set & Meaning & Unit & YAML key \\ \midrule
$d_E$ coeff. & Egg development rate & $1/^\circ\text{C}^2, 1/^\circ\text{C}, 1$ & \texttt{egg\_dev\_a,b,c} \\
$d_L$ coeff. & Larva development rate & $1/^\circ\text{C}^2, 1/^\circ\text{C}, 1$ & \texttt{larva\_dev\_a,b,c} \\
$d_P$ coeff. & Pupa development rate & $1/^\circ\text{C}^2, 1/^\circ\text{C}, 1$ & \texttt{pupa\_dev\_a,b,c} \\
$S_E$ & Egg survival (amp, mean, sigma) & dim., $^\circ\text{C}, ^\circ\text{C}$ & \texttt{egg\_survival\_...} \\
$S_L$ & Larva survival (amp, mean, sigma) & dim., $^\circ\text{C}, ^\circ\text{C}$ & \texttt{larva\_survival\_...} \\
$S_P$ & Pupa survival (amp, mean, sigma) & dim., $^\circ\text{C}, ^\circ\text{C}$ & \texttt{pupa\_survival\_...} \\
$\mu_A$ coeff. & Adult mortality & $1/^\circ\text{C}^2, 1/^\circ\text{C}, 1$ & \texttt{adult\_mort\_a,b,c} \\
$F$ coeff. & Fecundity & $a_F, b_F, T_{\max}, c_F$ & \texttt{fecundity\_a,b,Tmax,c} \\
$K$ & Host scaling factor & dimensionless & (derived) \\ \bottomrule
\end{tabular}
\end{table}


\paragraph{Rule engine and spatial registry.}  
Behavioural rules (e.g., “\texttt{feed}”, “\texttt{lay\_eggs}”, “\texttt{rest\_in\_building}”, etc.) are defined in a YAML 
configuration file and evaluated using a JavaScript engine (GraalVM). The spatial registry implements a lazy grid with a 
configurable cell size (default $0.001/^\circ\text{C} \approx 110 m$). Neighbour queries are performed by checking cells 
within a Chebyshev distance $\lceil \text{radius}/\text{cellSize}\rceil$ and cache results for one second to reduce 
computational cost.


\subsubsection{Simulation Configuration}
% how to configure a simulation
\par
All simulation parameters are defined in a single YAML file (\texttt{simulation.yaml}). The file specifies:  
\par
\begin{itemize}
  \item \textbf{Time settings}: start date, total number of ticks, tick duration (15-minutes).  
  \item \textbf{File paths}: study area shapefile, elevation, building density, population rasters, and the climate NetCDF.  
  \item \textbf{Species parameters}: coefficients for development, survival, fecundity, and adult mortality.  
  \item \textbf{Tokens}: mapping between environment variable names (e.g., “temperature”) and raster layers.  
  \item \textbf{Agent layers and rules}: each layer contains a list of rules with a condition, an action, and a priority.  
  \item \textbf{Seeding}: number of breeding sites and mosquitoes, habitat grid resolution, and whether to seed uniformly or 
  using a weighted habitat suitability map.  
  \item \textbf{Project defaults}: agent search radius, step size, maximum age.
\end{itemize}
\par
At startup, the \texttt{App} class loads the configuration, initialises the time manager, loads rasters and climate data, 
creates the spatial registry, builds agent layers, seeds the initial population, and starts the simulation engine.

\subsubsection{Result Analysis}
% details on the analysis notebook how are simulation result analyzed details
Simulation outputs are exported as CSV snapshots every 100 ticks (25 days). Each snapshot contains per-agent records 
with tick count, agent ID, layer, type, coordinates, alive flag, age, stage, energy, gravid status, and for inert agents, egg/larval 
counts and water volume. Environmental values (temperature, precipitation, etc) are attached to each agent's row via bilinear interpolation 
from the current climate frame.
\par
The analysis notebook (Python) performs the following steps:
\par
\begin{enumerate}
  \item \textbf{Daily aggregation}: The CSV is read in chunks (100 000 rows) to compute daily adult, larva, and pupa counts, as well 
  as tank statistics (mean water volume, total eggs/larvae), using the tick-to-day conversion ($96$ ticks per day).  
  \item \textbf{Population dynamics}: Stage counts are plotted, and stage durations are estimated from cross-correlation of smoothed 
  population peaks. Theoretical durations are computed from species parameters at the mean observed temperature.  
  \item \textbf{Environmental correlation}: Pearson correlation between daily adult counts and mean temperature / precipitation is calculated.  
  \item \textbf{Suitability mapping}: A kernel density estimate (KDE) of adult points (bandwidth 0.002° ≈ 220 m) is generated and exported 
  as a GeoTIFF. A breeding site potential raster is created by rasterising the sum of eggs and larvae in water tanks on the last tick.  
  \item \textbf{Validation}: The continuous Boyce index \citep{2024.EA.IAC} and the proportion of occurrence points falling into the top 20\% 
  of the suitability surface are computed using independently collected occurrence data (from GBIF, iNaturalist, Literature, clipped to 
  the study area).  
  \item \textbf{Export}: Daily summaries and a sample of raw data are saved for further analysis.
\end{enumerate}
\par
All figures and rasters are stored in a user-defined output directory.

\subsection{Case Study Design}
\label{sec:case_study}
\subsubsection{Vector and Study Site}

\subsubsection{Vector Biology}

\subsubsection{Data Sources}

\subsubsection{Simulation Assupmtions, Rules and Configuration}


\section{Results}
\label{sec:results}


\section{Discussion}
\label{sec:discussion}


\section{Conclusion}
\label{sec:conclusion}


\section*{Acknowledgments}
\par
This research was supported by the International Center of Insect Physiology and Ecology (ICIPE), 
the University of KwaZulu-Natal, and the International Institute of Tropical Agriculture (IITA). 
We acknowledge the providers of the open data used in this study, including GBIF, the Copernicus 
Climate Change Service (ERA5-Land), the Malaria Atlas Project, and the Vector Atlas consortium. 
We particularly thank the data contributors and curators who make their work publicly available, 
enabling research of this nature.

\section*{Code availability}

\bibliographystyle{apalike}
\bibliography{bibliography.bib}

\end{document}